---
title: "HW1"
author: "Yudong Feng"
date: "Sep.14 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Problem 1:
```{r}
library(tidyverse)
library(nycflights13)
```

(a) The type of variables is

```{r}
names(diamonds)
```

(b) The data dimensions are

```{r}
dim(diamonds)
```

(c) To find the most expensive diamond,

```{r}
diamonds_df <- as.data.frame(diamonds)
sort(diamonds_df[,"price"],decreasing = TRUE)[1:10]
```
the price is 18823, so we can do
```{r}
diamonds_exp = diamonds_df[diamonds_df["price"]==18823,]
```
the characteristics of the most expensive diamond is
```{r}
diamonds_exp
```

(d) The "Premium" diamonds are

```{r}
diamonds_prem = diamonds_df[diamonds_df["cut"]=="Premium",]
```
the number of "Premium" diamonds is 13791
```{r}
dim(diamonds_prem)
```


Problem 2
We use ggplot2 package to analyze relationships between price and cut, clarity, color, and carat.
```{r}
library(ggplot2)
```

By common sense, carat seems related to price. To analyze the relationship, we can do 
```{r}
ggplot(data = diamonds) + geom_point(mapping = aes(x = carat, y = price))
```
It shows that carat is highly related to price, the more carats the diamond has, the higher the price is.
Now we look at the color,
```{r}
ggplot(data = diamonds) + geom_smooth(mapping = aes(x = carat, y = price,color = color))
```
It seems like when carat < 2, color is related to price.
Then for clarity,
```{r}
ggplot(data = diamonds) + geom_smooth(mapping = aes(x = carat, y = price,color = clarity))
```
It shows that clarity is related to price, IF>VVS1>VVS2>VS1>VS2>SI1>SI2>I1.
Similarly, for cut,
```{r}
ggplot(data = diamonds) + geom_smooth(mapping = aes(x = carat, y = price, color = cut))
```
When carat < 3, good, very good, premium and ideal diamonds have similar prices, but fair diamonds have a lower price. When 3<carat<4, the cut has limited influence on price. When carat > 4, due to rareness of big diamonds, most of these diamonds only have fair cut.

Problem 3
The function is 
```{r}
my.bisect <- function(fun,interval,tol){
  a = interval[1]
  b = interval[2]
  while (b-a>tol){
    if (fun(a)*fun((a+b)/2) <= 0){
      b = (a+b)/2
    }else{
      a = (a+b)/2
    }
  }
  a
}
```

(a)
g and derivative of g are
```{r}
g <- function(x){
  log(x)/(1+x)
}

dg <- function(x){
  (1+x^(-1)-log(x))/((1+x)^2)
}
```
The root of dg = 0 is 
```{r}
my.bisect(dg,c(1,5),0.001)
```

The maximium of g is 
```{r}
g(my.bisect(dg,c(1,5),0.001))
```

(b)
Define my.bisect.hm to return the number of iterations,
```{r}
my.bisect.hm <- function(fun,interval,tol){
  a = interval[1]
  b = interval[2]
  k = 0
  while (b-a>tol){
    k = k + 1
    if (fun(a)*fun((a+b)/2) <= 0){
      b = (a+b)/2
    }else{
      a = (a+b)/2
    }
  }
  k
}
```
The number of iterations s.t. error<0.001,
```{r}
my.bisect.hm(dg,c(1,5),0.001)
```

(c) For tol to 0.001
```{r}
my.bisect.hm(dg,c(1,5),0.001)
```
compared to uniroot function,
```{r}
uniroot(dg,c(1,5),tol = 0.001)
```

For tol to 0.0001,
```{r}
my.bisect.hm(dg,c(1,5),0.0001)
```
compared to uniroot function,
```{r}
uniroot(dg,c(1,5),tol = 0.0001)
```

The uniroot function has less number of iterations.

(d)
```{r}
k_seq = c(0)*25
tol=0.0001
for (k in 4:25){
  c = 0
  d = k^0.5
  while (d-c>tol){
    mid = (c+d)/2
    fun.c = pt((k*c^2/(k+1-c^2))^0.5,k) - pt(((k-1)*c^2/(k-c^2))^0.5,k-1)
    fun.mid = pt((k*mid^2/(k+1-mid^2))^0.5,k) - pt(((k-1)*mid^2/(k-mid^2))^0.5,k-1)
    if (fun.c*fun.mid < 0){
      d = (c+d)/2
    }else{
      c = (c+d)/2
    }
  }
  k_seq[k]=d
}

k = 100 
  c = 0
  d = k^0.5
  while (d-c>tol){
    mid = (c+d)/2
    fun.c = pt((k*c^2/(k+1-c^2))^0.5,k) - pt(((k-1)*c^2/(k-c^2))^0.5,k-1)
    fun.mid = pt((k*mid^2/(k+1-mid^2))^0.5,k) - pt(((k-1)*mid^2/(k-mid^2))^0.5,k-1)
    if (fun.c*fun.mid < 0){
      d = (c+d)/2
    }else{
      c = (c+d)/2
    }
  }
k_seq[1]=d

k = 500 
  c = 0
  d = k^0.5
  while (d-c>tol){
    mid = (c+d)/2
    fun.c = pt((k*c^2/(k+1-c^2))^0.5,k) - pt(((k-1)*c^2/(k-c^2))^0.5,k-1)
    fun.mid = pt((k*mid^2/(k+1-mid^2))^0.5,k) - pt(((k-1)*mid^2/(k-mid^2))^0.5,k-1)
    if (fun.c*fun.mid < 0){
      d = (c+d)/2
    }else{
      c = (c+d)/2
    }
  }
k_seq[2]=d

k = 1000 
  c = 0
  d = k^0.5
  while (d-c>tol){
    mid = (c+d)/2
    fun.c = pt((k*c^2/(k+1-c^2))^0.5,k) - pt(((k-1)*c^2/(k-c^2))^0.5,k-1)
    fun.mid = pt((k*mid^2/(k+1-mid^2))^0.5,k) - pt(((k-1)*mid^2/(k-mid^2))^0.5,k-1)
    if (fun.c*fun.mid < 0){
      d = (c+d)/2
    }else{
      c = (c+d)/2
    }
  }
k_seq[3]=d

k_seq
```
The first three items are k=100, 500, 1000, and the rest are k=4:25.

Problem 4
The function is 
```{r}
cauchycdf <- function(x,eta,theta){
  delta_x = seq(from=min(-abs(x),eta)-1000*theta,to=x,by=0.001)
  d = (((delta_x-eta)/theta)^2+1)*theta*pi
  0.001*sum(d^(-1))
}
```
Compare to pcauchy,
```{r}
cauchycdf(0,0,1)
pcauchy(0,0,1)
```

```{r}
cauchycdf(9,-1,5)
pcauchy(9,-1,5)
```

```{r}
cauchycdf(-11,2,12)
pcauchy(-11,2,12)
```